---
title: "Double Interval Censoring for General Delays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Double Interval Censoring for General Delays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk-opts, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,  # Stan models are too heavy to run in CI; code is shown for reference
  purl     = FALSE
)
```

## Overview

The package's existing **type 5** data format handles delays where the
**exposure** time is interval-censored (known only within a window) but the
**event** time is observed exactly. This is the typical setting for incubation
period estimation: the date of symptom onset is known but the precise moment of
infection is not.

For delays other than the incubation period -- such as the **serial interval**
(symptom onset to symptom onset), **generation time** (infection to infection),
or **time from onset to hospitalisation** -- *both* endpoints may be
interval-censored. This vignette demonstrates the **type 6** data format, which
handles this general double-censoring case correctly.

## The problem with the naive approach

A tempting shortcut is to form the outer delay bounds
$[\,O_L - E_R,\; O_R - E_L\,]$ and treat the result as a standard
interval-censored delay (type 5). This is equivalent to assigning uniform
probability to all delays in that outer interval. However, the correct
weight function is **trapezoidal**: delays near the centre of the outer
interval are more consistent with the observed windows than delays near the
edges. Using outer bounds therefore over-disperses the likelihood and biases
parameter estimates, with bias growing as censoring windows widen relative to
the mean delay.

The type 6 likelihood evaluates the correct 1-D integral

$$
L \;\propto\; \int_{E_L}^{E_R}
  \bigl[F_D(O_R - e) - F_D(O_L - e)\bigr]\, de
$$

via Gauss-Legendre quadrature, which requires only CDF evaluations and adds
negligible computational cost.

## Setup

```{r setup, purl=FALSE}
library(ddsynth)
library(rstan)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)
library(patchwork)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Simulating doubly-censored delay data

We simulate a hypothetical study of the **serial interval** -- the time between
symptom onset in a primary case and symptom onset in a secondary case. The true
distribution is log-normal with population mean $\mu_0 = \log(6)$ and
log-standard-deviation $\phi = 0.45$ (mean serial interval approximately 7
days).

Censoring structure:

- **Exposure window** (`expo_lower`, `expo_upper`): symptom onset of the
  primary case is known only to within a 1-3 day window (e.g. recalled
  approximately).
- **Event window** (`event_lower`, `event_upper`): symptom onset of the
  secondary case is recorded to the nearest day, yielding a 1-day interval.

```{r simulate, purl=FALSE}
simulate_double_censored <- function(n,
                                     mu    = log(6),
                                     sigma = 0.45,
                                     seed  = 101) {
  set.seed(seed)

  # True serial intervals
  D_true <- rlnorm(n, meanlog = mu, sdlog = sigma)

  # Exposure window: onset of primary case known within 1-3 days
  expo_width  <- sample(1:3, n, replace = TRUE)
  expo_lower  <- floor(runif(n, 0, 20))            # arbitrary calendar anchor
  expo_upper  <- expo_lower + expo_width

  # True exposure time: uniform within window
  E_true <- runif(n, expo_lower, expo_upper)

  # True event time and daily-rounded event window
  O_true      <- E_true + D_true
  event_lower <- floor(O_true)
  event_upper <- event_lower + 1L

  # Ensure positive delay bounds (drop any negative outer-bound rows)
  naive_lower <- event_lower - expo_upper
  naive_upper <- event_upper - expo_lower
  keep        <- naive_lower > 0

  data.frame(
    expo_lower  = expo_lower[keep],
    expo_upper  = expo_upper[keep],
    event_lower = event_lower[keep],
    event_upper = event_upper[keep],
    naive_lower = naive_lower[keep],
    naive_upper = naive_upper[keep],
    D_true      = D_true[keep]
  )
}

dat <- simulate_double_censored(n = 200)

cat("Simulated", nrow(dat), "observations\n")
cat("True delay summary (days):\n")
print(summary(dat$D_true))
```

## Formatting data for the model

### Type 6: correct double-censored likelihood

Each observation contributes one row with its four bounds. Identical window
combinations are collapsed into a frequency table to reduce the size of the
Stan data object.

```{r format-type6, purl=FALSE}
dat_freq6 <- dat |>
  group_by(expo_lower, expo_upper, event_lower, event_upper) |>
  summarise(freq_count = n(), .groups = "drop")

dataset_dc <- list(
  expo_lower  = dat_freq6$expo_lower,
  expo_upper  = dat_freq6$expo_upper,
  event_lower = dat_freq6$event_lower,
  event_upper = dat_freq6$event_upper,
  freq_count  = as.integer(dat_freq6$freq_count)
)

cat("Type 6 frequency table: ", nrow(dat_freq6), "distinct interval combinations\n")
```

### Type 5 (naive): outer-bounds approximation

For comparison, we also prepare the same data using the naive approach: each
observation's outer delay bounds $[O_L - E_R,\; O_R - E_L]$ are passed as a
type 5 interval-censored dataset.

```{r format-type5, purl=FALSE}
dat_freq5 <- dat |>
  group_by(naive_lower, naive_upper) |>
  summarise(freq_count = n(), .groups = "drop")

dataset_naive <- list(
  freq_lower = dat_freq5$naive_lower,
  freq_upper = dat_freq5$naive_upper,
  freq_count = as.integer(dat_freq5$freq_count)
)

cat("Type 5 (naive) frequency table:", nrow(dat_freq5), "distinct intervals\n")
```

## Preparing Stan data and compiling the model

```{r compile, purl=FALSE}
stan_model_code <- compile_stan_model(model = "factorised")
```

```{r stan-data, purl=FALSE}
# True parameter values for reference
true_mu    <- log(6)
true_sigma <- 0.45

# --- Type 6 (correct) ---
stan_data_dc <- prepare_stan_data_from_datasets(
  datasets   = list(serial_interval = dataset_dc),
  dist_type  = 1,   # log-normal
  custom_priors = list(
    mu0_sd       = 1.0,
    log_tau_mean = 0.0,
    log_tau_sd   = 0.5,
    log_phi_mean = -0.7,
    log_phi_sd   = 0.5
  )
)

# --- Type 5 naive ---
stan_data_naive <- prepare_stan_data_from_datasets(
  datasets   = list(serial_interval = dataset_naive),
  dist_type  = 1,
  custom_priors = list(
    mu0_sd       = 1.0,
    log_tau_mean = 0.0,
    log_tau_sd   = 0.5,
    log_phi_mean = -0.7,
    log_phi_sd   = 0.5
  )
)
```

## Fitting both models

```{r fit, purl=FALSE}
sampling_args <- list(
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  thin    = 1,
  control = list(adapt_delta = 0.9, max_treedepth = 10),
  seed    = 42
)

cat("Fitting type 6 (double-censored) model...\n")
fit_dc <- do.call(
  rstan::sampling,
  c(list(object = stan_model_code, data = stan_data_dc), sampling_args)
)

cat("Fitting type 5 (naive outer-bounds) model...\n")
fit_naive <- do.call(
  rstan::sampling,
  c(list(object = stan_model_code, data = stan_data_naive), sampling_args)
)
```

## Convergence diagnostics

```{r diagnostics, purl=FALSE}
check_convergence <- function(fit, label) {
  s     <- summary(fit)$summary
  rhat  <- s[, "Rhat"]
  n_eff <- s[, "n_eff"]
  cat(sprintf(
    "[%s] Max Rhat: %.3f | Min n_eff: %.0f | Rhat > 1.05: %d\n",
    label,
    max(rhat,  na.rm = TRUE),
    min(n_eff, na.rm = TRUE),
    sum(rhat > 1.05, na.rm = TRUE)
  ))
}

check_convergence(fit_dc,    "Type 6 (correct)")
check_convergence(fit_naive, "Type 5 (naive)  ")
```

## Parameter recovery

We compare posterior estimates of $\mu_0$ (population log-mean) and
$\phi$ (log-standard-deviation) against the true simulation values.

```{r recovery, purl=FALSE}
extract_summary <- function(fit, label) {
  sims <- rstan::extract(fit)
  data.frame(
    model     = label,
    parameter = c("mu0 (log-mean)", "phi (log-SD)"),
    true_val  = c(true_mu, true_sigma),
    post_mean = c(mean(sims$mu0),       mean(sims$phi)),
    post_med  = c(median(sims$mu0),     median(sims$phi)),
    ci_low    = c(quantile(sims$mu0, 0.025), quantile(sims$phi, 0.025)),
    ci_high   = c(quantile(sims$mu0, 0.975), quantile(sims$phi, 0.975))
  )
}

recovery_df <- bind_rows(
  extract_summary(fit_dc,    "Type 6 (correct)"),
  extract_summary(fit_naive, "Type 5 (naive)")
)

print(recovery_df, digits = 3)
```

The type 6 model should recover both parameters well. The naive type 5 model
will tend to over-estimate $\phi$ (attributing the extra dispersion introduced
by the wide outer bounds to the delay distribution itself) and may show a small
bias in $\mu_0$.

## Visualisation

### Posterior distributions for key parameters

```{r plot-posteriors, fig.width=9, fig.height=4, purl=FALSE}
draws_dc    <- rstan::extract(fit_dc)
draws_naive <- rstan::extract(fit_naive)

posterior_df <- bind_rows(
  data.frame(model = "Type 6 (correct)", mu0 = draws_dc$mu0,    phi = draws_dc$phi),
  data.frame(model = "Type 5 (naive)",   mu0 = draws_naive$mu0, phi = draws_naive$phi)
)

p_mu0 <- ggplot(posterior_df, aes(x = mu0, fill = model, colour = model)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  geom_vline(xintercept = true_mu, linetype = "dashed", linewidth = 0.8) +
  scale_fill_npg()   +
  scale_colour_npg() +
  labs(title    = expression(paste("Posterior: ", mu[0], " (log-mean)")),
       subtitle = "Dashed line = true value",
       x        = expression(mu[0]),
       y        = "Density",
       fill     = NULL, colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_phi <- ggplot(posterior_df, aes(x = phi, fill = model, colour = model)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  geom_vline(xintercept = true_sigma, linetype = "dashed", linewidth = 0.8) +
  scale_fill_npg()   +
  scale_colour_npg() +
  labs(title    = expression(paste("Posterior: ", phi, " (log-SD)")),
       subtitle = "Dashed line = true value",
       x        = expression(phi),
       y        = "Density",
       fill     = NULL, colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_mu0 + p_phi + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Posterior predictive CDFs

```{r plot-cdfs, fig.width=8, fig.height=5, purl=FALSE}
x_seq <- seq(0.1, 25, length.out = 400)

cdf_dc    <- compute_predictive_cdf(fit_dc,    "lognormal", x_seq = x_seq, n_draws = 500)
cdf_naive <- compute_predictive_cdf(fit_naive, "lognormal", x_seq = x_seq, n_draws = 500)

# True CDF for comparison
true_cdf_df <- data.frame(
  x   = x_seq,
  cdf = plnorm(x_seq, meanlog = true_mu, sdlog = true_sigma)
)

cdf_plot_df <- bind_rows(
  cdf_dc$summary    |> mutate(model = "Type 6 (correct)"),
  cdf_naive$summary |> mutate(model = "Type 5 (naive)")
)

model_colours <- c("Type 6 (correct)" = "#0072B5", "Type 5 (naive)" = "#BC3C29")

ggplot() +
  geom_ribbon(
    data = cdf_plot_df,
    aes(x = x, ymin = low, ymax = high, fill = model),
    alpha = 0.20
  ) +
  geom_line(
    data = cdf_plot_df,
    aes(x = x, y = median, colour = model),
    linewidth = 1.0
  ) +
  geom_line(
    data = true_cdf_df,
    aes(x = x, y = cdf),
    colour = "black", linetype = "dashed", linewidth = 0.9
  ) +
  scale_fill_manual(values   = model_colours) +
  scale_colour_manual(values = model_colours) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1)) +
  labs(
    title    = "Posterior predictive CDFs: type 6 vs naive type 5",
    subtitle = "Dashed black line = true log-normal CDF | Shaded bands = 95% credible intervals",
    x        = "Delay (days)",
    y        = "Cumulative probability",
    fill     = NULL, colour   = NULL
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

### Illustrating the naive over-dispersion

The plot below shows the distribution of observed delay intervals under the two
approaches for a random sample of 50 observations. The outer-bounds approach
produces wider, overlapping intervals that inflate apparent delay uncertainty.

```{r plot-intervals, fig.width=8, fig.height=5, purl=FALSE}
set.seed(7)
idx <- sample(nrow(dat), 50)
samp <- dat[idx, ] |>
  arrange(D_true) |>
  mutate(obs = row_number())

interval_df <- bind_rows(
  samp |> transmute(obs, approach = "Type 6 (correct)",
                    lower = event_lower - expo_upper,
                    upper = event_upper - expo_lower,
                    mid   = (lower + upper) / 2),
  samp |> transmute(obs, approach = "Type 5 (naive)",
                    lower = naive_lower, upper = naive_upper,
                    mid   = (lower + upper) / 2)
)

ggplot(interval_df, aes(y = obs, colour = approach)) +
  geom_segment(aes(x = lower, xend = upper, yend = obs),
               linewidth = 0.6, alpha = 0.7) +
  geom_point(data = samp |> mutate(obs = row_number()),
             aes(x = D_true, y = obs),
             colour = "black", shape = 4, size = 1.5, inherit.aes = FALSE) +
  scale_colour_manual(values = model_colours) +
  labs(
    title    = "Observed delay intervals for 50 sampled cases",
    subtitle = "Crosses mark the true simulated delay",
    x        = "Delay (days)",
    y        = "Observation (sorted by true delay)",
    colour   = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank())
```

## When to use type 6 vs type 5

| Situation | Recommended format |
|---|---|
| Event time is recorded to the exact day, exposure window is 0-1 days wide | Type 5 (or type 4) |
| Event time is exact; exposure window spans several days | Type 5 |
| Both endpoints interval-censored; windows narrow relative to the mean delay | Type 5 with outer bounds is acceptable |
| Both endpoints interval-censored; windows wide (e.g. > 30% of mean delay) | **Type 6** |
| Serial interval, generation time, or time-to-event with uncertain start | **Type 6** |

As a rule of thumb, if the sum of the exposure and event window widths
exceeds roughly a third of the expected delay, the naive outer-bounds
approximation will produce a meaningfully biased estimate of $\phi$ and
should be replaced by type 6.

## Session Info

```{r session-info, eval=TRUE, purl=FALSE}
sessionInfo()
```
