---
title: "Introduction to delaydistributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to delaydistributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(delaydistributions)
```

## Overview

The **delaydistributions** package provides three complementary approaches for
synthesising a delay distribution from summary statistics of previously observed
delays:

| Function | Method |
|---|---|
| `fit_gamma_mom()`, `fit_lognormal_mom()`, `fit_weibull_mom()` | Method of moments |
| `mle_from_summary()` | Maximum likelihood from summary statistics |
| `bayesian_synthesis()` | Bayesian hierarchical synthesis |

All methods take the **mean** and either the **variance** or **standard
deviation** of previously observed delays as inputs.  Sample sizes (`n`) can
optionally be supplied so that studies with more observations contribute more to
the final estimate.

---

## Method of moments

The method-of-moments estimators are closed-form expressions that match the
distributional mean and variance to the observed values.

```{r mom}
# Fit a Gamma distribution to a mean of 5 days and variance of 4 days^2
fit_g <- fit_gamma_mom(mean = 5, variance = 4)
fit_g

# Fit a Log-Normal distribution using the SD instead of the variance
fit_ln <- fit_lognormal_mom(mean = 5, sd = 2)
fit_ln

# Fit a Weibull distribution (solved numerically)
fit_w <- fit_weibull_mom(mean = 5, variance = 4)
fit_w
```

---

## Maximum likelihood from summary statistics

When summary statistics from multiple studies are available, `mle_from_summary()`
combines them by numerically maximising a log-likelihood that is constructed
from the observed means and variances.

```{r mle}
fit_mle <- mle_from_summary(
  mean         = c(5.0, 5.2, 4.8),
  sd           = c(2.0, 1.8, 2.1),
  n            = c(100L, 80L, 120L),
  distribution = "gamma"
)
fit_mle$parameters
fit_mle$loglik
```

---

## Bayesian synthesis

`bayesian_synthesis()` uses a conjugate Normal-Gamma model on the log scale to
produce a full posterior distribution over the delay parameters.  The result
includes a data frame of posterior samples that can be used for downstream
uncertainty quantification.

```{r bayesian}
set.seed(42)
result <- bayesian_synthesis(
  mean = c(5.0, 5.2, 4.8),
  sd   = c(2.0, 1.8, 2.1),
  n    = c(100L, 80L, 120L)
)

# Posterior summary
cat("Posterior mean of meanlog:", result$posterior_mean_meanlog, "\n")
cat("Posterior mean of sdlog:  ", result$posterior_mean_sdlog, "\n")

# First few posterior samples
head(result$posterior_samples)
```
