---
title: "Real-Time Line List Analysis with Right Truncation Correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-Time Line List Analysis with Right Truncation Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk-opts, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE,
  purl     = FALSE
)
```

## Overview

During an active outbreak, a line list collects cases as they are reported.
Cases with long delays between exposure and symptom onset are less likely to
have been reported by the analysis date -- they appear to be absent, not
because they don't exist but because they haven't happened yet or haven't been
recorded. This is **right truncation**: the observed sample is a biased subset
of all cases, enriched for short delays.

A second complication is **double interval censoring**: both the exposure date
(e.g. start of a contact window) and the onset date are often known only to the
day, not the exact hour.

The **type 7** data format in `ddsynth` handles both simultaneously:

- Doubly interval-censored delays (both exposure and onset known to the day).
- Right truncation correction: only cases with onset before the analysis date
  $T$ are observed; the denominator of the likelihood accounts for this.
- Right-censored cases: individuals with a known exposure window but no onset
  recorded yet contribute the survival function $1 - F_D(T - x_i)$.

## The likelihood

Let $e_i \sim \text{Uniform}(E^L_i, E^R_i)$ be the latent exposure time and
$s_i \sim \text{Uniform}(S^L_i, S^R_i)$ the latent onset time. Marginalising
over both latent times gives:

**Observed cases** (onset recorded, $s_i < T$):

$$
L_i \;\propto\; \frac{
  \int_{E^L_i}^{E^R_i} \bigl[F_D(S^R_i - e) - F_D(S^L_i - e)\bigr]\, de
}{
  \int_{E^L_i}^{E^R_i} F_D(T - e)\, de
}
$$

**Right-censored cases** (exposure known, onset not yet seen):

$$
L_i \;\propto\; \int_{E^L_i}^{E^R_i} \bigl[1 - F_D(T - e)\bigr]\, de
$$

Both integrals are evaluated via 7-point Gauss-Legendre quadrature, requiring
only repeated CDF evaluations and adding negligible overhead over type 6.

## Setup

```{r setup, purl=FALSE}
library(ddsynth)
library(rstan)
library(dplyr)
library(ggplot2)
library(patchwork)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Simulating a line list

We simulate an outbreak of 300 cases. The true incubation period is
log-normal with $\mu_0 = \log(6)$ and $\phi = 0.45$ (mean $\approx$ 7 days).
The analysis date is day 40 from outbreak start. Cases with exposures close to
day 40 may not yet have developed symptoms -- these become right-censored
observations.

```{r simulate, purl=FALSE}
set.seed(42)

reference_date <- as.Date("2024-01-01")
T_cal          <- as.Date("2024-02-10")   # analysis date
T_num          <- as.numeric(T_cal - reference_date)  # calendar day 40

n_cases  <- 300
true_mu0 <- log(6)
true_phi <- 0.45

# Exposure start uniform over outbreak period; 1-day window
expo_start_num <- runif(n_cases, 0, 35)
expo_end_num   <- expo_start_num + 1

# True latent exposure and incubation delay
expo_true  <- runif(n_cases, expo_start_num, expo_end_num)
delay_true <- rlnorm(n_cases, meanlog = true_mu0, sdlog = true_phi)
onset_true <- expo_true + delay_true

# Only cases with onset before T are observed
observed_mask <- onset_true < T_num

# For observed cases: onset known to the day (floor to 1-day window)
onset_lower <- floor(onset_true)
onset_upper <- onset_lower + 1

# Build line list
linelist <- data.frame(
  exposure_start = reference_date + floor(expo_start_num),
  exposure_end   = reference_date + floor(expo_start_num) + 1,
  onset_lower    = ifelse(observed_mask, reference_date + onset_lower, NA),
  onset_upper    = ifelse(observed_mask, reference_date + onset_lower + 1, NA)
)
linelist$onset_lower <- as.Date(linelist$onset_lower, origin = "1970-01-01")
linelist$onset_upper <- as.Date(linelist$onset_upper, origin = "1970-01-01")

cat("Total cases in line list:", nrow(linelist), "\n")
cat("Observed (onset recorded):", sum(observed_mask), "\n")
cat("Right-censored (no onset yet):", sum(!observed_mask), "\n")
```

## Preparing the dataset

Convert dates to numeric (days from `reference_date`). All bounds in a type 7
dataset must share the same calendar origin.

```{r prepare-data, purl=FALSE}
to_num <- function(d) as.numeric(d - reference_date)

dataset_ll <- list(
  expo_lower      = to_num(linelist$exposure_start),
  expo_upper      = to_num(linelist$exposure_end),
  event_lower     = ifelse(!is.na(linelist$onset_lower),
                           to_num(linelist$onset_lower), NA_real_),
  event_upper     = ifelse(!is.na(linelist$onset_upper),
                           to_num(linelist$onset_upper), NA_real_),
  freq_count      = rep(1L, nrow(linelist)),
  truncation_time = T_num
)

# Sanity checks
stopifnot(all(dataset_ll$event_upper[!is.na(dataset_ll$event_upper)] <= T_num))
stopifnot(all(dataset_ll$expo_upper < T_num))
```

The table below summarises the fields:

| Field | Description |
|---|---|
| `expo_lower`, `expo_upper` | Exposure window (calendar days from reference) |
| `event_lower`, `event_upper` | Onset window; `NA` for right-censored rows |
| `freq_count` | Number of cases with this combination |
| `truncation_time` | Analysis date $T$ (calendar days from reference) |

## Compiling the model and preparing Stan data

```{r compile, purl=FALSE}
stan_model_code <- compile_stan_model(model = "factorised")
```

```{r stan-data, purl=FALSE}
stan_data_ll <- prepare_stan_data_from_datasets(
  datasets  = list(linelist = dataset_ll),
  dist_type = 1,   # log-normal
  custom_priors = list(
    mu0_mean     = log(5),
    mu0_sd       = 1.0,
    log_tau_mean = 0.0,
    log_tau_sd   = 0.5,
    log_phi_mean = -0.7,
    log_phi_sd   = 0.5
  )
)
```

## Fitting the model

```{r fit, purl=FALSE}
fit_ll <- rstan::sampling(
  stan_model_code,
  data    = stan_data_ll,
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  control = list(adapt_delta = 0.9, max_treedepth = 10),
  seed    = 42
)

print(fit_ll, pars = c("mu0", "phi"))
```

## Summarising the posterior

```{r posterior-summary, purl=FALSE}
draws <- rstan::extract(fit_ll)

mean_delay <- exp(draws$mu0 + draws$phi^2 / 2)

cat(sprintf(
  "Mean incubation period: %.1f days (95%% CI: %.1f -- %.1f)\n",
  median(mean_delay),
  quantile(mean_delay, 0.025),
  quantile(mean_delay, 0.975)
))
cat("True mean:", round(exp(true_mu0 + true_phi^2 / 2), 1), "days\n")
```

## Temporal trend diagnostics

Even with truncation correction, it is important to verify that the apparent
shortening of delays near the analysis date is fully accounted for by the
model. Two diagnostics address this.

**Panel A -- Truncation envelope**: Individual observed delays are plotted
against onset date. The coloured curves show model-predicted quantiles of the
delay distribution *truncated* at the maximum observable delay for that onset
date ($T - \text{exposure midpoint}$). If the observed delays track the
envelope, truncation alone explains the pattern. If delays fall systematically
below the envelope, there is genuine shortening beyond what truncation predicts.

**Panel B -- PIT over time**: The probability integral transform value
$F_D(d_i) / F_D(T - x_i)$ should be approximately Uniform(0, 1) with no
temporal trend if truncation is the only source of apparent delay shortening.
A downward loess trend reveals a genuine delay reduction over time.

```{r temporal-diagnostics, purl=FALSE}
posterior <- extract_posterior_draws(fit_ll, stan_data_ll)

plot_temporal_trend(
  dataset        = dataset_ll,
  posterior      = posterior,
  reference_date = reference_date,
  probs          = c(0.5, 0.75, 0.95),
  n_draws        = 200L
)
```

### Interpreting the plots

| Pattern | Interpretation |
|---|---|
| Points track envelope; PIT flat and centred | Truncation fully explains apparent shortening |
| Points below envelope; PIT trends downward | Genuine delay reduction -- consider a time-varying model |
| Points above envelope; PIT trends upward | Delays genuinely increasing or reporting lag |
| PIT multimodal or U-shaped | Parametric family mis-specified |

## Combining with historic data

A major advantage of the `ddsynth` framework is the ability to pool line list
estimates with estimates from earlier studies that used different data formats.
Six hypothetical historic studies are defined below, mixing three summary types:

- **Type 1** (median + range): Studies C and F
- **Type 2** (median + IQR): Studies A, B, and E
- **Type 3** (mean + SD): Study D

All values are consistent with a log-normal distribution with phi (log-SD)
around 0.45 but slightly varying population means, reflecting genuine
between-study heterogeneity. This is important: `phi` is a **shared**
parameter across all datasets, so summary statistics must be compatible with
the same distributional shape.

With 7 datasets total (6 historic + 1 line list), the between-study
heterogeneity parameter `tau` becomes identifiable from data, and the
predicted quantities are averaged over the full predictive distribution.

```{r combine-historic, purl=FALSE}
# --- Type 2: median + IQR ---
# Study A: log-normal(log(5.5), 0.45)
dataset_A <- list(n = 60L, median = 5.50, Q1 = 4.06, Q3 = 7.45)

# Study B: log-normal(log(6.5), 0.45)
dataset_B <- list(n = 80L, median = 6.50, Q1 = 4.80, Q3 = 8.81)

# Study E: log-normal(log(6.2), 0.45)
dataset_E <- list(n = 120L, median = 6.20, Q1 = 4.58, Q3 = 8.40)

# --- Type 1: median + range (min/max as 1st and nth order statistics) ---
# Study C: log-normal(log(7.0), 0.45), n=45
dataset_C <- list(n = 45L, median = 7.00, min = 2.82, max = 17.37)

# Study F: log-normal(log(5.8), 0.45), n=35
dataset_F <- list(n = 35L, median = 5.80, min = 2.45, max = 13.73)

# --- Type 3: mean + SD ---
# Study D: log-normal(log(6.0), 0.45) -> mean = 6.64, SD = 3.15
dataset_D <- list(n = 100L, mean = 6.64, sd = 3.15)

stan_data_joint <- prepare_stan_data_from_datasets(
  datasets = list(
    study_A  = dataset_A,
    study_B  = dataset_B,
    study_C  = dataset_C,
    study_D  = dataset_D,
    study_E  = dataset_E,
    study_F  = dataset_F,
    linelist = dataset_ll
  ),
  dist_type = 1,
  custom_priors = list(
    mu0_mean     = log(6),
    mu0_sd       = 1.0,
    log_tau_mean = -1.0,   # prior centred on tau ~ 0.37; realistic between-study SD
    log_tau_sd   = 0.5,
    log_phi_mean = -0.7,
    log_phi_sd   = 0.5
  )
)

fit_joint <- rstan::sampling(
  stan_model_code,
  data    = stan_data_joint,
  chains  = 4,
  iter    = 4000,
  warmup  = 1000,
  control = list(adapt_delta = 0.95, max_treedepth = 12),
  seed    = 42
)

print(fit_joint, pars = c("mu0", "phi", "tau"))
```

With more than 5 datasets, `tau` is identifiable from data and the reported
`pred_mean`, `pred_median` etc. are averaged over the predictive distribution
for a new study, incorporating both within-study uncertainty and
between-study heterogeneity. The type 7 likelihood for the line list
automatically corrects for truncation and right-censoring.
