---
title: "Simulation Study: Evaluating the Hierarchical Bayesian Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study: Evaluating the Hierarchical Bayesian Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	collapse = TRUE,
	comment = "#>"
)
```

## Overview

This vignette evaluates the performance of the hierarchical Bayesian model across
a wide range of simulated scenarios. The study examines how well the model
recovers the true population parameters ($\mu_0$, $\tau$, $\phi$) when individual
observations are unavailable and only summary statistics are reported.

Three standard parametric families are considered — **log-normal**, **Gamma**, and
**Weibull** — along with two extended families: **Burr XII** and **Generalised
Gamma** (GG). Five summary-statistic types are evaluated:

| `summary_type` | Fields reported |
|---|---|
| 1 | Median + range (min, max) |
| 2 | Median + IQR (Q1, Q3) |
| 3 | Mean + SD |
| 4 | Frequency table (rounded counts) |
| 5 | Mixed (multiple types across studies) |

> **Note on Generalised Gamma identifiability.** The GG distribution has three
> parameters ($\mu_0$, $\phi$, $\kappa$). In a hierarchical synthesis setting,
> certain combinations of summary-statistic types and sample sizes render the
> shape parameter $\kappa$ weakly identified. Scenarios flagged by the
> `gg_heuristic` rule are excluded from the summaries below; only scenarios where
> GG was judged estimable are retained.

Performance is assessed via:

- **Credible-interval coverage** — empirical 95% CI coverage for model parameters
  ($\mu_0$, $\tau$, $\phi$) and for two predictive quantiles (P50, P95); nominal
  target: 0.95.
- **Median bias** — median of (posterior median $-$ true value) across replicates
  for parameters and predictive quantiles; target: 0.
- **Mean absolute error (MAE)** — for model parameters.
- **Integrated quadratic distance (IQD)** and **Weighted Interval Score (WIS)**
  — measures of predictive distribution accuracy; lower is better.
- **MCMC convergence rates** — proportion of replicates with $\hat{R} < 1.05$.

## Setup

```{r libs}
library(ddsynth)
library(tidyverse)
library(ggsci)
library(here)        # used only for inst/extdata path resolution below
```

## Load Pre-computed Results

Running the full simulation study is computationally intensive. The results used
here were generated by `analysis/new_simulation_study.R` and saved to disk. All
scenario metadata (distribution family, summary type, sample sizes, true
parameters) is embedded directly in the results file — no separate scenario
library is needed.

```{r load}
# Resolve path: prefer the installed-package location; fall back to the
# vignettes/ source file which is tracked in git and available during
# pkgdown / R CMD build vignette builds.
rds_path <- system.file("extdata", "simulation_results_all.rds", package = "ddsynth")
if (!nzchar(rds_path)) {
  rds_path <- file.path(here(), "vignettes", "simulation_results_all.rds")
}
has_results <- file.exists(rds_path)
if (!has_results) {
  message("simulation_results_all.rds not found; skipping analysis chunks.")
} else {
  res_out <- readRDS(rds_path)
  cat(sprintf("Loaded %d simulation rows across %d unique scenarios.\n",
              nrow(res_out), length(unique(res_out$scenario_name))))
  cat(sprintf("Distribution families: %s\n",
              paste(sort(unique(res_out$dist_type)), collapse = ", ")))
  n_skipped <- sum(!is.na(res_out$skipped_reason))
  if (n_skipped > 0L)
    cat(sprintf("Note: %d rows excluded as non-identifiable (skipped_reason = '%s').\n",
                n_skipped, unique(na.omit(res_out$skipped_reason))))
}
```

## Summarise Results

`create_results_summary()` aggregates per-replicate outputs, restricting to
converged, estimable runs. The resulting table contains, for each scenario:

- **Coverage** — empirical 95% CI coverage for $\mu_0$, $\tau$, $\phi$, and
  the predictive median (P50) and 95th percentile (P95).
- **Median bias** — median of (posterior median $-$ true value) for parameters
  and predictive quantiles.
- **MAE** — mean absolute error for parameters.
- **IQD** and **WIS** — mean predictive scoring rules across replicates.

```{r summary, eval = has_results}
summary_res <- create_results_summary(res_out)
cat(sprintf("Summary covers %d scenario-level rows.\n", nrow(summary_res)))
```

## Coverage of 95% Credible Intervals — Model Parameters

Each point represents one scenario. Points near the dashed red line (0.95)
indicate well-calibrated uncertainty quantification. Points are coloured by the
number of datasets and shaped by the within-study sample size.

```{r coverage, fig.width=9, fig.height=8, eval = has_results}
create_coverage_plot(summary_res)
```

## Median Bias — Model Parameters

The dashed red line at zero indicates unbiased estimation. Systematic departures
from zero suggest that certain data configurations (e.g., very small $n$ or
sparse summary types) introduce recoverable or persistent bias.

```{r bias, fig.width=9, fig.height=8, eval = has_results}
create_bias_plot(summary_res)
```

## Mean Absolute Error — Model Parameters

MAE captures the magnitude of estimation error irrespective of direction.
Smaller values across larger datasets and richer summary types confirm that
pooling information across studies reduces estimation uncertainty.

```{r mae, fig.width=9, fig.height=8, eval = has_results}
create_mae_plot(summary_res)
```

## Coverage of 95% Credible Intervals — Predictive Median (P50)

Coverage for the posterior predictive median. Points near the dashed red line
(0.95) indicate that the credible intervals reliably contain the true population
median. Points are coloured by number of datasets and shaped by within-study
sample size.

```{r pred_median_coverage, fig.width=9, fig.height=4, eval = has_results}
create_pred_median_coverage_plot(summary_res)
```

## Coverage of 95% Credible Intervals — Predictive 95th Percentile (P95)

Coverage for the posterior predictive 95th percentile. Tail quantiles are
typically harder to estimate and may show lower coverage in data-sparse
scenarios.

```{r pred_q95_coverage, fig.width=9, fig.height=4, eval = has_results}
create_pred_q95_coverage_plot(summary_res)
```

## Median Bias — Predictive Median (P50)

Systematic over- or under-estimation of the predictive median across scenarios.
Departures from zero can reflect inadequate pooling in sparse conditions.

```{r pred_median_bias, fig.width=9, fig.height=4, eval = has_results}
create_pred_median_bias_plot(summary_res)
```

## Median Bias — Predictive 95th Percentile (P95)

Bias in the posterior predictive 95th percentile. Persistent over-estimation of
the tail is common when few datasets are available or the summary type provides
limited information about the upper tail.

```{r pred_q95_bias, fig.width=9, fig.height=4, eval = has_results}
create_pred_q95_bias_plot(summary_res)
```

## Integrated Quadratic Distance (IQD)

IQD measures the average squared discrepancy between the true and estimated
predictive CDFs across the full support. Lower values indicate that the posterior
predictive distribution closely matches the data-generating distribution.

```{r iqd, fig.width=9, fig.height=4, eval = has_results}
create_iqd_plot(summary_res)
```

## Weighted Interval Score (WIS)

WIS is a proper scoring rule that penalises both sharpness (narrow intervals)
and miscoverage. Lower values indicate better overall predictive accuracy.
Results here mirror the IQD analysis, providing a complementary view of
predictive performance.

```{r wis, fig.width=9, fig.height=4, eval = has_results}
create_wis_plot(summary_res)
```

## Summary Table

The table below aggregates all metrics across scenarios within each
distribution-family × summary-type cell (averaging over number of datasets and
within-study sample size). Coverage values are shown as percentages (nominal
target: 95%). Predictive bias is the median signed error in days (target: 0).
IQD and WIS are means over scenarios (lower = better).

```{r summary_table, eval = has_results}
library(kableExtra)

# Shared footnote text reused by both tables
.fn <- paste0(
  "Coverage: empirical proportion of replicates where the true value fell ",
  "inside the 95% posterior credible interval. ",
  "Predictive bias: median signed error (posterior median estimate \u2212 true value) across replicates. ",
  "IQD: integrated quadratic distance between true and estimated predictive CDFs. ",
  "WIS: weighted interval score. ",
  "Gen. Gamma scenarios with fewer than 10 identifiable replicates were excluded prior to aggregation."
)

# Helper: format a summarised data frame into a kable
.fmt <- function(dat) {
  dat %>%
    transmute(
      `mu0`   = sprintf("%.1f", cov_mu0  * 100),
      `tau`   = sprintf("%.1f", cov_tau  * 100),
      `phi`   = sprintf("%.1f", cov_phi  * 100),
      `P50`   = sprintf("%.1f", cov_p50  * 100),
      `P95`   = sprintf("%.1f", cov_p95  * 100),
      `P50 `  = sprintf("%+.2f", bias_p50),
      `P95 `  = sprintf("%+.2f", bias_p95),
      `IQD`   = sprintf("%.3f", iqd),
      `WIS`   = sprintf("%.2f", wis)
    )
}

# ── Table 1: distribution × summary type ─────────────────────────────────────
t1_dat <- summary_res %>%
  mutate(
    dist_label = factor(dist_type,
      levels = c("lognormal", "gamma", "weibull", "burr12", "gengamma"),
      labels = c("Log-normal", "Gamma", "Weibull", "Burr XII", "Gen. Gamma")),
    st_label = factor(summary_type, levels = 1:5,
      labels = c("Median+Range", "Median+IQR", "Mean+SD", "Freq Table", "Mixed"))
  ) %>%
  group_by(dist_label, st_label) %>%
  summarise(
    n_scen   = n(),
    cov_mu0  = mean(coverage_mu0,         na.rm = TRUE),
    cov_tau  = mean(coverage_tau,         na.rm = TRUE),
    cov_phi  = mean(coverage_phi,         na.rm = TRUE),
    cov_p50  = mean(coverage_pred_median, na.rm = TRUE),
    cov_p95  = mean(coverage_pred_q95,    na.rm = TRUE),
    bias_p50 = median(bias_pred_median,   na.rm = TRUE),
    bias_p95 = median(bias_pred_q95,      na.rm = TRUE),
    iqd      = mean(mean_iqd,             na.rm = TRUE),
    wis      = mean(mean_wis,             na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  arrange(dist_label, st_label)

tbl1 <- bind_cols(
  t1_dat %>% transmute(
    Distribution  = as.character(dist_label),
    `Summary type` = as.character(st_label),
    `N` = n_scen
  ),
  .fmt(t1_dat)
)

kbl(tbl1,
    col.names = c("Distribution", "Summary type", "N",
                  "\u03bc\u2080", "\u03c4", "\u03c6",
                  "P50", "P95", "P50", "P95", "IQD", "WIS"),
    align   = c("l", "l", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r"),
    caption = "Performance summary by distribution family and summary type") %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                full_width = TRUE, font_size = 12) %>%
  add_header_above(c(" " = 3,
                     "Parameter coverage (%)" = 3,
                     "Predictive coverage (%)" = 2,
                     "Predictive bias (days)" = 2,
                     "Scoring rules" = 2)) %>%
  pack_rows(index = t1_dat %>% count(dist_label) %>% deframe()) %>%
  footnote(general = .fn, general_title = "Note: ")
```

## Summary Table: By Data Availability

The table below pools results across distribution families and shows how
performance varies with the number of datasets and within-study sample size,
broken down by summary type. Rows are grouped by summary type.

```{r summary_table2, eval = has_results}
t2_dat <- summary_res %>%
  mutate(
    st_label = factor(summary_type, levels = 1:5,
      labels = c("Median+Range", "Median+IQR", "Mean+SD", "Freq Table", "Mixed")),
    n_datasets_bucket = case_when(
      n_datasets < 10  ~ "<10",
      n_datasets < 20  ~ "<20",
      n_datasets < 30  ~ "<30",
      n_datasets >= 30 ~ "30+"
    ) %>% factor(levels = c("<10", "<20", "<30", "30+")),
    n_obs_bucket = case_when(
      n_obs == 5  ~ "5",
      n_obs == 10 ~ "10",
      n_obs == 20 ~ "20",
      n_obs > 25  ~ "25+"
    ) %>% factor(levels = c("5", "10", "20", "25+"))
  ) %>%
  group_by(st_label, n_datasets_bucket, n_obs_bucket) %>%
  summarise(
    n_scen   = n(),
    cov_mu0  = mean(coverage_mu0,         na.rm = TRUE),
    cov_tau  = mean(coverage_tau,         na.rm = TRUE),
    cov_phi  = mean(coverage_phi,         na.rm = TRUE),
    cov_p50  = mean(coverage_pred_median, na.rm = TRUE),
    cov_p95  = mean(coverage_pred_q95,    na.rm = TRUE),
    bias_p50 = median(bias_pred_median,   na.rm = TRUE),
    bias_p95 = median(bias_pred_q95,      na.rm = TRUE),
    iqd      = mean(mean_iqd,             na.rm = TRUE),
    wis      = mean(mean_wis,             na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  arrange(st_label, n_datasets_bucket, n_obs_bucket)

tbl2 <- bind_cols(
  t2_dat %>% transmute(
    `Summary type` = as.character(st_label),
    `N datasets`   = as.character(n_datasets_bucket),
    `N obs`        = as.character(n_obs_bucket),
    `N`            = n_scen
  ),
  .fmt(t2_dat)
)

kbl(tbl2,
    col.names = c("Summary type", "N datasets", "N obs", "N",
                  "\u03bc\u2080", "\u03c4", "\u03c6",
                  "P50", "P95", "P50", "P95", "IQD", "WIS"),
    align   = c("l", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r", "r"),
    caption = "Performance summary by summary type, number of datasets, and within-study sample size (pooled over distribution families)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                full_width = TRUE, font_size = 12) %>%
  add_header_above(c(" " = 4,
                     "Parameter coverage (%)" = 3,
                     "Predictive coverage (%)" = 2,
                     "Predictive bias (days)" = 2,
                     "Scoring rules" = 2)) %>%
  pack_rows(index = t2_dat %>% count(st_label) %>% deframe()) %>%
  footnote(general = .fn, general_title = "Note: ")
```

## MCMC Convergence Rates

The convergence plot shows the proportion of simulation replicates where all
chains reached $\hat{R} < 1.05$ and the effective sample size was adequate.
Values at the dashed red line (1.0) indicate perfect convergence across all
replicates. Lower rates in sparse conditions (few datasets, small $n$) highlight
where the model may require more informative priors or additional iterations.
Scenarios deemed non-identifiable (e.g. certain GG configurations) are excluded.

```{r convergence, fig.width=9, fig.height=8, eval = has_results}
create_convergence_plot(res_out)
```
