---
title: "Simulation Study: Evaluating the Hierarchical Bayesian Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study: Evaluating the Hierarchical Bayesian Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Overview

This vignette evaluates the performance of the hierarchical Bayesian model across
a wide range of simulated scenarios. The study examines how well the model
recovers the true population parameters ($\mu_0$, $\tau$, $\phi$) when individual
observations are unavailable and only summary statistics are reported.

Three parametric families are considered — **log-normal**, **Gamma**, and
**Weibull** — along with five summary-statistic types:

| `summary_type` | Fields reported |
|---|---|
| 1 | Median + range (min, max) |
| 2 | Median + IQR (Q1, Q3) |
| 3 | Mean + SD |
| 4 | Frequency table (rounded counts) |
| 5 | Mixed (multiple types across studies) |

Performance is assessed via **credible-interval coverage**, **median bias**,
**mean absolute error (MAE)**, and **MCMC convergence rates**, each broken down
by distribution family, summary type, number of datasets, and within-study
sample size.

## Setup

```{r libs}
library(ddsynth)
library(tidyverse)
library(ggsci)
```

## Load Pre-computed Results

Running the full simulation study is computationally intensive. The results used
here were generated by `run_simulation_study_generalized()` and saved to disk.

```{r load}
rds_path    <- system.file("extdata", "simulation_results_all.rds", package = "ddsynth")
has_results <- nzchar(rds_path) && file.exists(rds_path)
if (!has_results) {
  message("simulation_results_all.rds not found; skipping analysis chunks.")
} else {
  res_out <- readRDS(rds_path)
}
```

## Build the Scenario Library

`generate_scenario_library()` constructs the full factorial grid of scenarios
used in the study, covering homogeneous, mixed, varied-$n$, and frequency-table
conditions.

```{r scenarios}
cat("Generating scenario library...\n")

scenarios <- as_tibble(generate_scenario_library(
  include_homogeneous   = TRUE,
  include_mixed         = TRUE,
  include_varied_n      = TRUE,
  include_freq_table    = TRUE,
  include_burr12        = TRUE,
  include_gengamma      = TRUE,
  include_gg_limitation = TRUE
))

cat("Generated", nrow(scenarios), "scenarios\n")
```

## Summarise Results

`create_results_summary()` aggregates the per-replicate outputs (restricting to
converged runs) and joins the scenario metadata. The resulting table contains,
for each scenario:

- **Coverage** — empirical 95% credible-interval coverage for $\mu_0$, $\tau$,
  and $\phi$ (nominal target: 0.95).
- **Median bias** — median of (posterior median $-$ true value) across replicates
  (target: 0).
- **MAE** — mean absolute error across replicates.
- **IQD** — mean integrated quadratic distance between the true and estimated
  predictive distributions.

```{r summary, eval = has_results}
summary_res <- create_results_summary(res_out, scenarios) %>%
  mutate(n_obs = n_obs_mean)
```

## Coverage of 95% Credible Intervals

Each point represents one scenario. Points near the dashed red line (0.95)
indicate well-calibrated uncertainty quantification. Points are coloured by the
number of datasets and shaped by the within-study sample size.

```{r coverage, fig.width=9, fig.height=8, eval = has_results}
create_coverage_plot(summary_res)
```

## Median Bias of Parameter Estimates

The dashed red line at zero indicates unbiased estimation. Systematic departures
from zero suggest that certain data configurations (e.g., very small $n$ or
sparse summary types) introduce recoverable or persistent bias.

```{r bias, fig.width=9, fig.height=8, eval = has_results}
create_bias_plot(summary_res)
```

## Mean Absolute Error of Parameter Estimates

MAE captures the magnitude of estimation error irrespective of direction.
Smaller values across larger datasets and richer summary types confirm that
pooling information across studies reduces estimation uncertainty.

```{r mae, fig.width=9, fig.height=8, eval = has_results}
create_mae_plot(summary_res)
```

## MCMC Convergence Rates

The convergence plot shows the proportion of simulation replicates where all
chains reached $\hat{R} < 1.05$ and the effective sample size was adequate.
Values at the dashed red line (1.0) indicate perfect convergence across all
replicates. Lower rates in sparse conditions (few datasets, small $n$) highlight
where the model may require more informative priors or additional iterations.

```{r convergence, fig.width=9, fig.height=8, eval = has_results}
create_convergence_plot(res_out, scenarios)
```
